### 2019-03-13 manuscript/discussion.tex
- backup of mostly done first 2/3 Discussion sections - addition of refs and minor text changes in Results

%B:  at least two surprising outcomes: (1)  D and P on background of C, and (2) SC not as dead of an end as it was in goldberg2010, if P is partitioned out (can be framed as kind of path-dependence, i.e. qIC path different than rho path, or same as (1), so that I and C are less different on the background of D.
We investigated the individual and joint effects of ploidy and breeding system states in diversification of Solanaceae.
Most significantly, we found that the effects of ploidy, inferred as significant when it is the only trait under consideration, are overshadowed by hidden traits, or breeding system when added to models.  
Interestingly, the difference in diversification rates between the diploids and polyploids is negligible on the background of self-compatibility.
In the same vein, we recovered a smaller difference in diversification rates between self-incompatible and self-compatible lineages against the diploid genomic background.
%Inference of transition rates can appreciably across 
Motivated by the widespread findings of paleopolyploid ancestry across angiosperms, we also examined and found modest support for diploidization in the history of the family. %B: "mixed" "not strong" what's the right word here?  E: not "weak", maybe "not strong" or "not nearly as strong as anything else"
Disentangling the patterns of diversification linked to multiple traits is difficult, but our strategy allows tests of intricate hypotheses regarding a heterogenous diversification process.
Below, we discuss these key findings and highlight the most interesting and paradoxical implications.
We additionally provide an outline for broader application of phylogenetic comparative methods in the context of interactions among traits and diversification process.

\subsection*{Disentangling trait effects on diversification}

% confounding effects of other traits

% - value of approach
Disentangling the patterns of diversification linked to  multiple traits allowed us the opportunity to model and test complex hypotheses about the heterogenous diversification process in Solanaceae.
Species are created and go extinct based on multiple and often highly correlated phenotypes. 
However, estimating the speciation and extinction rates remains challenging. 
By carefully aggregating and curating ploidy and breeding system data for Solanaceae, we showed how  the complex interaction between ploidy and breeding system  can lead to diversification conclusions that differ from the diversification conclusions of models that focus on one trait at a time.

% - what happens if PD only, Hidden states
Previous analyses of polyploidy linked to diversification showed how diploids had greater net diversification rates than polyploids when exploring across multiple clades of the angiosperm phylogeny \citet{mayrose_2011, mayrose_2015}. 
Using the most complete dataset for ploidy in a phylogenetic tree for Solanaceae, we were able to replicate the results found by \citet{mayrose_2011} where polyploids have a slower net diversification compare to diploids.
However, we expanded this study to accommodate  background heterogeneity in the diversification process.
When adding heterogeneity we found that it was more likely that an unobserved trait linked to diploid state was the one leading the net diversification patterns, and that there were some ``second-class"  diploids that were not different from  polyploids  in diversification terms (Figure 2A).
This result made us question whether there was a second trait linked to diploidy that could make a differential in the diversification progress.

%     main finding about our traits
For Solanaceae, looking into self-incompatibility was a logical step.
Previous studies have shown that self-incompatible Solanaceae species have also higher rates of diversification compared to their self-compatible counterparts \citep{goldberg_2012}. 
Again, in our sample when looking only into breeding system linked to the diversification process we were able to replicate the slower net diversification for self-compatible taxa (Figure 2C). When this model was expanded to incorporate hidden states we found again that some self-incompatible taxa have faster net diversification rates than self-compatible taxa. However, there is again a  different class of taxa where those differences are not as striking (Figure 2D).  However, investigating both  both polyploidy and breeding system simultaneously for every species in our sample, allowed us to further investigate why some diploids were quantitatively different than polyploids, or why some self-incompatible taxa would be different in diversification terms from some self-compatible when a second class is not.
In the three-state diversification model ID/CD/CP, we represented the complex interactions between ploidy and breeding system. We found that self-incompatible diploids have faster and positive rates than self-compatible diploids and polyploids, and that the difference between the  rates of net diversification of self-compatible diploids is not as large (Figure 2E) as first found by binary trait diversification models (Figure 2A).
This result is key, it not only aligned with the net diversification results from the D/P+A/B model, where a `	`hidden-trait" seem to be dictating the diversification pattern but it also aligned with the hidden state model I/C+A/B where some class of self-compatible net diversification rate might be overlapping with two net diversification rates of self-incompatible.
% Removed by R. Therefore, we consider that finding a heterogenous result in the hidden trait approaches should be be treated as evidence of a second trait that is necessary to consider.
% Removed by R. Pursuing knowledge of  such trait can result on a clearer picture of the importance of trait linked diversification patterns, but also on a better reconstruction on past events in phylogenies.

@@

We find that, in Solanaceae, a diploid lineage is much more likely to take a
one-step  ID$\rightarrow$CP pathway to polyploidy rather than a two-step ID$\rightarrow$CD$\rightarrow$CP pathway.
This result is significant because it has a number of implications for the inference of history of breeding and sexual systems in flowering plants.
First, support for the direct pathway lends credence to the idea that breakdown of SI by WGD---by generation of diploid `heteroalleleic' pollen---may often trigger the evolution of gender dimorphism across angiosperms \citep{miller_2000,roberston_2011}.

%missing deeper discussion of with/without SDD/SID estimates and comparison with Robertson et al. 2011.

% qualifying paragraph for pathways
% "This finding is not without a hitch..."
We implicitly modeled trait changes in both of these pathways as anagenetic, transitions occurring within lineages, without associated speciation events.
However, losses of SI (I$\rightarrow$C transitions) are often associated with cladogenesis in Solanaceae \citep{goldberg_2012}, and polyploidization (D$\rightarrow$P transition) also seem to co-occur with speciation events in a number of other families \citep{zhan_2016,freyman_2017}.
Models that fail to consider such cladogenetic changes may make misleading estimates of anagenetic transition rates.
In the only available data bearing on this problem, \citet{goldberg_2012} found that extending the *SSE framework to allow cladogenetic changes did not substantially affect inference of net diversification rates.
Nevertheless, a distinct possibility remains that allowing for cladogenetic transitions could tip the scales further in favor of the one-step pathway to SC polyploidy (CP), but this remains to be tested in future work.
Specifically, the multi/hidden state SSE-based framework we employed here could be extended to incorporate cladogenetic transitions.
% We could also list %W+%R idea above, for stochastic mapping?
Another useful line of inquiry could examine the path-dependent diversification rates, asking whether diversification rates of polyploid SC lineages differ depending on the path taken.

% And, are diversification rates of polyploid SC lineages different if one or other path was taken? \citet{charlesworth1985} % E: This is a great question.  We don't answer it with this round of analyses.  But discuss?
% E: To elaborate: In the IC/CD/CP model, the CP state would be subdivided into two hidden values, corresponding to the influx from the two pathways.  Those two states, CP_X and CP_Y, could have different lambda and mu.

%W: 3) I'll bet allowing cladogenetic transitions would make the one-step ID/CP even more dominant.
%    I added the paragraph below regarding cladogenetic transitions, but it doesn't really fit with
%    the rest of the discussion, so modify or discard as needed:
% E: A small defense of excluding clado is that it didn't affect the diversification results in goldberg2010 vs goldberg2012

@@

%R: By adding diploidization to models of polyploidy linked to diversification it is possible to recover this complicated scenario and to reconcile genomic evidence with stochastic models. 
%In the cases where ploidy level was assigned by us based chromosome multiplicity at the genus level, estimating diploidization  might be a potential issue due to ploidy misclassification.

@@

Genomic evidence indicates that a paleo-polyploidization may have occurred prior to the origin of the family Solanaceae, rendering all current `diploids' secondarily derived. 
First, \citet{ku2000} and \citet{blanc2004} posited that the lineage leading to tomato, \textit{Solanum lycopersicum}, may have experienced one or more WGDs.
A subsequent analysis of synteny between grape and \textit{Solanum} genomes, as well as the distribution between inferred paralogs within genomes of \textit{Solanum} (tomato and potato) each suggested that this lineage experienced a likely round of ancient genome duplication or triplication \citep{tomato2012}. 
The age of the peak of paralog Ks distances, is approximately 71 million years \citep{tomato2012}. 
If this is the case, then all of the genomes may have been subsequently re-diploidized, yielding the common extant base chromosome numbers in this and related families, n=11-12, considered diploid \citep{robertson_2011}. 
%Not unlikely, because Arabidopsis n=5.

@@

% PATHWAYS -- mutational opportunity vs lineage transition rate reveals something about selection 
%
% Starting with SI-D, selection favoring establishment of SC-D seems more common than that of SC-P.
% This is because the mutation rates are likely higher to SC-P, but rate 3x lower. "
%Husband2012: The prevalence of polyploidy within plants has led to debate about the influence of genome duplication on net species diversification (Otto and Whitton 2000). Researchers have argued, on one hand, that polyploidy may simply accumulate as a unidirectional mutation with little influence on the rate or trajectory of evolutionary divergence (Stebbins 1971; Meyers and Levin 2006). This idea is based on the premise that ploidy is more likely to increase than decrease in time. Although reversals do occur, the resulting haploids are rela- tively rare and often of reduced fitness and high sterility (Stebbins 1980; Ramsey and Schemske 2002). Alternatively, polyploidy itself may enhance evolutionary diversification leading to increased divergence and species richness (Otto and Whitton 2000). Elevated diversification may occur if polyploidy accelerates adaptive divergence and evolution of reproductive isolation (Otto and Whitton 2000) or reduces species extinction. 
%The presence of intraspecific ploidy diversity has three main causes. First, genome duplication is an ongoing genomic mutation within many natural populations and ploidy polymorphism is therefore expected when poly- ploid formation is common. In fact, Ramsey and Schemske (1998) estimated that polyploids form at the rate of 10 5 per individual per generation, well above the genic mutation rate, and in some cases it may be even higher than that (Husband 2004). Second, depending on the frequency and pathways of polyploid formation, diploids and polyploids may be maintained as genetically homogeneous populations, which are justifiably treated as single taxonomic species. Finally, ploidy polymorphism may arise as an artefact of our imperfect taxonomic approach. Since morphological divergence is often used as a criterion for species delinea- tion, polyploid lineages may not be recognized at the species level, especially in autopolyploids, which often have a close resemblance to their diploid progenitors (Soltis et al. 2007).

>>>>

%B:  at least two surprising outcomes: (1)  D and P on background of C, and (2) SC not as dead of an end as it was in goldberg2010, if P is partitioned out (can be framed as kind of path-dependence, i.e. qIC path different than rho path, or same as (1), so that I and C are less different on the background of D.
%
Although species are composed of vast assemblages of variable traits, many traits are heritable, and they may affect the propensity of species to perish and multiply \citep{lewontin_1970}, examining such macroevolutionary effects of many traits is rare, and remains challenging.
We investigated the influence of two traits, ploidy and breeding system states, in diversification of Solanaceae.
Most significantly, we found that the effects of ploidy, inferred as significant when it is the only trait under consideration, are overshadowed by hidden traits, or breeding system when added to models.  %R: added hidden trait.
Interestingly, the difference in diversification rates between the diploids and polyploids is negligible on the background of self-compatibility.
In the same vein, we recovered a smaller difference in diversification rates between self-incompatible and self-compatible lineages against the diploid genomic background.
Motivated by the widespread findings of paleopolyploid ancestry across angiosperms, we also examined and found modest support for diploidization in the history of the family. %B: how about "modest"? It's 3-ish, but "positive" seems like a really idiosyncratic usage to me, despite kass & raftery. was: %B: "mixed" "not strong" what's the right word here?  E: not "weak", maybe "not strong" or "not nearly as strong as anything else"
Disentangling the patterns of diversification linked to multiple traits is difficult, but our strategy allows tests of intricate hypotheses regarding a heterogenous diversification process.
Below, we discuss key findings and highlight the most interesting and paradoxical implications.
We additionally provide an outline for broader application of phylogenetic comparative methods in the context of interactions among traits and diversification process.

\subsection*{Disentangling trait effects on diversification} %B: Emma may not like "effects,"?

Previous analyses of the effects of ploidy on diversification showed that diploids had greater net diversification rates than polyploids when exploring across multiple clades of the angiosperm phylogeny \citet{mayrose_2011, mayrose_2015}. 
We obtain a broadly similar outcome, when considering only the effects of ploidy, but the results change significantly with joint consideration of hidden states, and breeding systems. 
The addition of a hidden trait results in blurring of differences in diversification rates between diploids and polyploids (Figure 2A). %B: fix fig ref
Simultaneous consideration of breeding system goes a long way in explaining variation in the inferred net diversification rates among ploidy states.
SI, which is itself associated with a high diversification rate, is the breeding system context in which many diploids occur.
Among SC species, diploids and polyploids are associated with similar diversification rates (Figure 2F). %B: fix fig ref
The best interpretation of these results is that some of the effects of a hidden trait seem to be dictated by  each of the two traits under consideration.
The finding is unique, and implications may be of wide interest, because it evaluates the relative influence of these two exceptionally interesting traits in a comparative phylogenetic framework.

Previous work has shown that SI species in Solanaceae are associated with higher diversification rates compared to their SC counterparts \citep{goldberg_2010}. 
When we inferred only the role of breeding system in the diversification process, we were able to replicate the slower net diversification for SC (Figure 2C).  %B: fix fig ref
Incorporating hidden states again resulted in SI associated with greater net diversification rates than SC in both hidden states (Figure 2D). 
However, the overall picture is a bit more complex.
Depending on the background hidden state, it is possible that a class of SC species may experience higher diversification rates than SI species.
Moreover, these SC lineages may not constitute a dead end (with negative diversification rates).

@@

% qualifying paragraph for pathways
We find that, in Solanaceae, SI diploid lineages are much more likely to take a
one-step  ID$\rightarrow$CP pathway to polyploidy rather than a two-step ID$\rightarrow$CD$\rightarrow$CP pathway.
This result is significant because it has a number of implications for the inference of history of breeding and sexual systems in flowering plants.
First, support for the direct pathway lends credence to the idea that breakdown of SI by WGD---by generation of diploid `heteroalleleic' pollen---may often trigger the evolution of gender dimorphism across angiosperms \citep{miller_2000}.
A more direct and definitive test of this hypothesis would additionally examine the propensity of polyploids generated through either pathway to become dioecious \citep{robertson_2011}.
% -- mutational opportunity vs lineage transition rate reveals something about selection 
Second, statistical phylogenetic approaches measure lineage transition rates.
Transitions depend on a combination of availability of individual mutation rates, which generate trait variation, and selection, which may cause disproportionate loss or fixation of these mutants.
Estimates of mutation rates are highly uncertain, but the total rate of breakdown of SI within diploids ranges is estimated to be on the order of  $\mu_{IC} = 10^{-5}$ per pollen grain, and includes breakdown by tetraploidization \citep{lewis1979}, which is itself estimated to occur on the order of $\mu_{DP} = 10^{-5}$ \citep{ramsey_1998}. 
Seemingly, then, the simple genic mutation rate that leads to loss of SI is at best equal to tetraploidization mutation rate, and possibly far lower.
We nevertheless recover transition rates from ID to CD ($q_{IC}$) threefold greater than $\rho_{I}$, indicating that selection restricts the fixation rate of new polyploids compared with new SC mutants \citep{robertson_2011}.

% qualifying paragraph for pathways
We implicitly modeled trait changes in both of these pathways as anagenetic, transitions occurring within lineages, without associated speciation events.
However, losses of SI (I$\rightarrow$C transitions) are often associated with cladogenesis in Solanaceae \citep{goldberg_2012}, and polyploidization (D$\rightarrow$P transition) also seem to co-occur with speciation events in a number of other families \citep{zhan_2016,freyman_2017}.
Models that fail to consider such cladogenetic changes may make misleading estimates of anagenetic transition rates.
In the only available data bearing on this problem, \citet{goldberg_2012} found that extending the *SSE framework to allow cladogenetic changes did not substantially affect inference of net diversification rates.
Nevertheless, a distinct possibility remains that allowing for cladogenetic transitions could tip the scales further in favor of the one-step pathway to SC polyploidy (CP), but this remains to be tested in future work. %B Will's comment alluded to, but not directly. 
Specifically, the multi/hidden state SSE-based framework we employed here could be extended to incorporate cladogenetic transitions. %B: We could also list %W+%R idea above, for stochastic mapping?
Another useful line of inquiry would examine the path-dependent diversification rates, asking whether diversification rates of polyploid SC lineages differ depending on whether they were generated via a one- or two-step pathway. %B: could elaborate if you see necessary. TODO
%B: also missing deeper discussion of with/without SDD/SID estimates and comparison with Robertson et al. 2011.

%B: And, are diversification rates of polyploid SC lineages different if one or other path was taken? % E: This is a great question.  We don't answer it with this round of analyses.  But discuss?
% E: To elaborate: In the IC/CD/CP model, the CP state would be subdivided into two hidden values, corresponding to the influx from the two pathways.  Those two states, CP_X and CP_Y, could have different lambda and mu.
%
%W: 3) I'll bet allowing cladogenetic transitions would make the one-step ID/CP even more dominant.
%    I added the paragraph below regarding cladogenetic transitions, but it doesn't really fit with
%    the rest of the discussion, so modify or discard as needed:
% E: A small defense of excluding clado is that it didn't affect the diversification results in goldberg2010 vs goldberg2012

@@

%B: TODO: add below to future goals.
%R: and I: polyploid lineage that has re-diploidized may enhance diversification.
%R: By adding diploidization to models of polyploidy linked to diversification it is possible to recover this complicated scenario and to reconcile genomic evidence with comparative phylogenetic models. 
%In the cases where ploidy level was assigned by us based chromosome multiplicity at the genus level, estimating diploidization  might be a potential issue due to ploidy misclassification.

@@

Genomic evidence indicates that a paleo-polyploidization may have occurred prior to the origin of the family Solanaceae, rendering all current `diploids' secondarily derived. 
First, \citet{ku2000} and \citet{blanc2004} posited that the lineage leading to tomato, \textit{Solanum lycopersicum}, may have experienced one or more WGDs.
A subsequent analysis of synteny between grape and \textit{Solanum} genomes, as well as the distribution between inferred paralogs within genomes of \textit{Solanum} (tomato and potato) each suggested that this lineage experienced a likely round of ancient genome duplication or triplication \citep{tomato2012}. 
The age of the peak of paralog Ks distances, is approximately 71 million years \citep{tomato2012}. 
If this is the case, then all of the genomes may have been subsequently re-diploidized, yielding the common extant base chromosome numbers in this and related families, n=11-12, presently considered to be diploid \citep{robertson_2011}. 
%Not unlikely, because Arabidopsis n=5.

@@

% PATHWAYS -- mutational opportunity vs lineage transition rate reveals something about selection 
%
% Starting with SI-D, selection favoring establishment of SC-D seems more common than that of SC-P.
% This is because the mutation rates are likely higher to SC-P, but rate 3x lower. "
%Husband2012: The prevalence of polyploidy within plants has led to debate about the influence of genome duplication on net species diversification (Otto and Whitton 2000). Researchers have argued, on one hand, that polyploidy may simply accumulate as a unidirectional mutation with little influence on the rate or trajectory of evolutionary divergence (Stebbins 1971; Meyers and Levin 2006). This idea is based on the premise that ploidy is more likely to increase than decrease in time. Although reversals do occur, the resulting haploids are relaively rare and often of reduced fitness and high sterility (Stebbins 1980; Ramsey and Schemske 2002). Alternatively, polyploidy itself may enhance evolutionary diversification leading to increased divergence and species richness (Otto and Whitton 2000). Elevated diversification may occur if polyploidy accelerates adaptive divergence and evolution of reproductive isolation (Otto and Whitton 2000) or reduces species extinction. 
%The presence of intraspecific ploidy diversity has three main causes. First, genome duplication is an ongoing genomic mutation within many natural populations and ploidy polymorphism is therefore expected when poly- ploid formation is common. In fact, Ramsey and Schemske (1998) estimated that polyploids form at the rate of 10 5 per individual per generation, well above the genic mutation rate, and in some cases it may be even higher than that (Husband 2004). Second, depending on the frequency and pathways of polyploid formation, diploids and polyploids may be maintained as genetically homogeneous populations, which are justifiably treated as single taxonomic species. Finally, ploidy polymorphism may arise as an artefact of our imperfect taxonomic approach. Since morphological divergence is often used as a criterion for species delinea- tion, polyploid lineages may not be recognized at the species level, especially in autopolyploids, which often have a close resemblance to their diploid progenitors (Soltis et al. 2007).

### 2019-03-13 manuscript/results.tex
- backup of mostly done first 2/3 Discussion sections - addition of refs and minor text changes in Results

Indeed, incorporation of lineage diversification rates changes conclusions regarding the relative importance of each pathway.
Beginning again with a single ID lineage, it is now necessary to consider its possible proliferation in the ID state, and also in the CD state for the two-step pathway, as well as to track the progression of the full resulting clade to the CP state.
In this situation, even over long timescales, the two-step pathway contributes less (\cref{figure:pathways}, right panels).
The lower rate of net diversification in the CD state, relative to ID, means that relatively fewer lineages are available to complete the second step of the two-step pathway, ending in CP.

>>>>

Indeed, incorporation of lineage diversification rates changes conclusions regarding the relative contribution of each pathway.
Beginning with a single ID lineage, assessment of relative pathway importance necessitates consideration of proliferation in the ID state, as well as the CD state for the two-step pathway, and to track the transitions to the CP state.
In this situation, even over long timescales, the two-step pathway contributes less to generation of CP (\cref{figure:pathways}, right panels).
The lower rate of net diversification in the CD state, relative to ID, means that relatively fewer lineages are available to complete the second step of the two-step pathway, ending in CP.
